---
title: "Interop with Bioconductor and Seurat Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interop with Bioconductor and Seurat Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows three common downstream workflows against a
singlify pipeline output: Bioconductor `scater` + `scran`, Seurat,
and RNA velocity via scVelo. The goal is to demonstrate that the
`singlify` package drops the pipeline outputs into standard ecosystem
objects without any reshape, rename, or dtype conversion.

## Setup

```{r setup}
library(singlify)

sample_dir <- "quant/scrna/GSE174/GSE174399/GSM5293863"
```

## Bioconductor: scater + scran

```{r scater}
library(SingleCellExperiment)
library(scater)
library(scran)

sce <- as_sce(sample_dir)

# Per-cell QC metrics are already in colData from the pipeline's
# cell_qc_metrics.tsv — but we can still run scater's own metrics for
# cross-validation.
sce <- scater::addPerCellQC(sce)

# Filter using the pre-loaded values (no manual mt_pct computation
# needed — the pipeline already did it).
keep <- sce$cell_qc_metrics__mt_pct < 20 &
        sce$cell_qc_metrics__total_umis > 500
sce <- sce[, keep]

# Size-factor normalization
sce <- scran::computeSumFactors(sce)
sce <- scater::logNormCounts(sce)

# HVG + PCA + UMAP
dec <- scran::modelGeneVarByPoisson(sce)
hvgs <- scran::getTopHVGs(dec, n = 2000)
sce <- scater::runPCA(sce, subset_row = hvgs)
sce <- scater::runUMAP(sce, dimred = "PCA")

# Clusters
g <- scran::buildSNNGraph(sce, k = 15, use.dimred = "PCA")
sce$cluster <- factor(igraph::cluster_louvain(g)$membership)

scater::plotUMAP(sce, colour_by = "cluster")
```

Note how the embedded metadata is still first-class:

```{r bioc_meta}
metadata(sce)$singlify[["gsm_id"]]
metadata(sce)$singlify[["protocol"]]
```

## Seurat: standard pipeline

```{r seurat}
library(Seurat)

obj <- as_seurat(sample_dir)

# Standard Seurat workflow — no renaming or reassay needed
obj <- subset(obj,
              subset = cell_qc_metrics__total_umis > 500 &
                       cell_qc_metrics__mt_pct < 20)

obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj, nfeatures = 2000)
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 30)
obj <- FindNeighbors(obj, dims = 1:30)
obj <- FindClusters(obj, resolution = 0.5)
obj <- RunUMAP(obj, dims = 1:30)

DimPlot(obj, group.by = "seurat_clusters")
```

GEO context is in `@misc$singlify`:

```{r seurat_meta}
obj@misc$singlify[["gsm_id"]]
obj@misc$singlify[["organism"]]
```

## RNA velocity

`as_seurat()` attaches `spliced` and `unspliced` as additional assays
with scvelo-compatible names. For the R `velociraptor` package
(Bioconductor wrapper around scVelo):

```{r velocity}
library(velociraptor)
library(SingleCellExperiment)

sce <- as_sce(sample_dir)

# velociraptor expects the spliced/unspliced assays
velo <- velociraptor::scvelo(sce)

# Or direct scVelo via reticulate if you prefer the Python API:
# scv <- reticulate::import("scvelo")
# adata <- scv$datasets$toy_data()  # replace with sce_to_anndata(sce)
# scv$tl$velocity(adata)
```

## Batch / cohort analyses

The per-sample outputs from a single-cell cohort can all be merged
because every one carries the GEO context in its metadata.

```{r cohort}
cohort_dirs <- list.dirs("quant/scrna/GSE174/GSE174399", recursive = FALSE)

sces <- lapply(cohort_dirs, as_sce)
names(sces) <- vapply(
    sces,
    function(x) metadata(x)$singlify[["gsm_id"]],
    character(1)
)

# Merge via cbind (for SCE)
common_genes <- Reduce(intersect, lapply(sces, rownames))
merged <- do.call(
    cbind,
    lapply(sces, function(x) x[common_genes, ])
)

# Each column now has a per-cell "sample" label if you carry it through
merged$sample <- rep(names(sces), vapply(sces, ncol, integer(1)))
```

## What's kept vs dropped

The singlify pipeline ships only non-redundant outputs. Matrices that
are element-wise derivable from others (`gene_counts = spliced +
unspliced + ambiguous`) and always-zero matrices on 3'/5' 10x and
drop-seq (`ambiguous`) are dropped from the NFS copy. Everything you
need for downstream analysis is still present:

| You need                  | File / assay              |
|---------------------------|---------------------------|
| Per-gene counts           | `counts` (the spliced-derived assay in SCE) |
| Velocity                  | `spliced` + `unspliced` assays |
| Alt splicing              | `sj_counts`, `splice_psi` |
| Per-exon / per-intron     | `altExp(sce, "exon_counts")` etc. |
| Mitochondrial heteroplasmy| `altExp(sce, "mt_heteroplasmy")` |
| VDJ usage                 | `altExp(sce, "vdj_gene_usage")` |
| EM-rescued counts         | `gene_counts_em` assay    |
| Per-cell QC               | `colData(sce)`            |
| Cell-cycle phase          | `colData(sce)$cell_cycle_scores__phase` |
| Doublet scores            | `colData(sce)$doublet_scores__doublet_score` |
| Sample / GEO context      | `metadata(sce)$singlify`  |
| Donor assignments         | read `donor_assignments.tsv` directly from `attr(dd, "path")` |
