---
title: "Quickstart: Reading Singlet Pipeline Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart: Reading Singlet Pipeline Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`singlet` reads the per-sample output directories produced by the
[singlet single-cell reprocessing pipeline](https://github.com/zdebruine/Singlet-AI)
— specifically the `.1pz` sparse-matrix files and the per-cell TSV
sidecars — into native R objects.

This vignette covers the three most common entry points:
`read_1pz()` (one file), `read_singlet_dir()` (a whole sample), and
`as_sce()` / `as_seurat()` (drop-in conversion to Bioconductor /
Seurat).

## Install

```{r install}
# From GitHub
install.packages("remotes")
remotes::install_github("zdebruine/Singlet-AI", subdir = "singlet/r")
```

You need a C++17 compiler and `libzstd` (>= 1.4) on the build machine.
On Linux: `apt install libzstd-dev` or `yum install libzstd-devel`.
On macOS: `brew install zstd`.

## Load

```{r load}
library(singlet)
```

## Read a single .1pz file

The lowest-level entry point takes a file path and returns a
`Matrix::dgCMatrix` whose attributes carry the embedded pipeline
metadata.

```{r read_one}
mat <- read_1pz("quant/scrna/GSE174/GSE174399/GSM5293863/gene_counts.1pz")

dim(mat)                           # 38606 x 16079
mat[1:5, 1:3]                      # top-left corner

# Embedded metadata — the GEO context the pipeline was launched with.
kv <- attr(mat, "user_kv")
kv[["gsm_id"]]                     # "GSM5293863"
kv[["gse_id"]]                     # "GSE174399"
kv[["organism"]]                   # "Homo sapiens"
kv[["protocol"]]                   # "10xv3"
kv[["singlet_version"]]           # e.g. "0.3.0"
kv[["pipeline_date"]]              # e.g. "2026-04-13"

# Writer's narrow-storage hint
attr(mat, "vt_code")               # 1 (uint8), 2 (uint16), or 3 (uint32)
```

## Read a whole pipeline output directory

`read_singlet_dir()` walks every `.1pz` file in the directory and
returns a named list. All the matrices share the same `user_kv`,
which is attached as an attribute on the list.

```{r read_dir}
dd <- read_singlet_dir("quant/scrna/GSE174/GSE174399/GSM5293863")
dd
# > singlet pipeline directory (11 matrices, gsm=GSM5293863)
# >   gene_counts          38606 x 16079, nnz=...
# >   spliced              38606 x 16079, nnz=...
# >   unspliced            38606 x 16079, nnz=...
# >   ambiguous            38606 x 16079, nnz=0
# >   exon_counts          310797 x 16079, nnz=...
# >   intron_counts        272191 x 16079, nnz=...
# >   sj_counts            ...
# >   splice_psi           ...
# >   mt_heteroplasmy      ...
# >   vdj_gene_usage       ...
# >   gene_counts_em       ...

names(dd)
dd$gene_counts                    # dgCMatrix, 38606 x 16079
dd$spliced                        # same shape
attr(dd, "user_kv")[["gsm_id"]]   # "GSM5293863"
```

Use `include =` or `exclude =` to load only specific matrices:

```{r read_dir_filter}
# Just the velocity trio
dd_vel <- read_singlet_dir(
    "quant/scrna/GSE174/GSE174399/GSM5293863",
    include = c("spliced", "unspliced", "ambiguous")
)
```

## Convert to SingleCellExperiment

```{r as_sce}
library(SingleCellExperiment)

sce <- as_sce("quant/scrna/GSE174/GSE174399/GSM5293863")
sce
# > class: SingleCellExperiment
# > dim: 38606 16079
# > assays(4): counts unspliced ambiguous gene_counts_em
# > altExpNames(6): exon_counts intron_counts sj_counts splice_psi mt_heteroplasmy vdj_gene_usage

# The embedded GEO metadata
metadata(sce)$singlet[["gsm_id"]]

# Per-cell QC metrics auto-loaded from the pipeline's cell_qc_metrics.tsv
head(colData(sce))
# > cell_qc_metrics__total_umis cell_qc_metrics__total_genes cell_qc_metrics__mt_pct ...

# Per-feature matrices live in altExp so they keep their own feature axis
altExp(sce, "exon_counts")        # 310797 x 16079
altExp(sce, "sj_counts")          # splice junctions
```

## Convert to Seurat

The Seurat adapter attaches the velocity trio as additional assays so
scvelo-compatible workflows run without extra renaming.

```{r as_seurat}
library(Seurat)

obj <- as_seurat("quant/scrna/GSE174/GSE174399/GSM5293863")
obj
# > An object of class Seurat
# > 77212 features across 16079 samples within 3 assays
# > Active assay: RNA (38606 features, ...)
# >  2 other assays present: unspliced, ambiguous

obj@misc$singlet[["gsm_id"]]

# Per-cell sidecar metadata in @meta.data (prefixed by the TSV source
# filename so different sidecars never collide)
head(obj@meta.data)
```

## What's in each matrix

The singlet pipeline writes multiple `.1pz` files per sample. A few
are **verified redundant** and are dropped from the NFS copy in recent
pipeline versions (see the pipeline's `quant/` manifest); the rest
are independent.

| File                  | Contents                                            | Derivable from |
|-----------------------|-----------------------------------------------------|----------------|
| `spliced.1pz`         | Per-gene spliced counts                             | —              |
| `unspliced.1pz`       | Per-gene unspliced counts (RNA velocity)            | —              |
| `ambiguous.1pz`       | Per-gene ambiguous reads                            | zero for 10x/drop-seq |
| `gene_counts.1pz`     | GeneFull = spliced + unspliced + ambiguous          | derivable (often dropped) |
| `exon_counts.1pz`     | **Per-exon** matrix (310797 features)               | —              |
| `intron_counts.1pz`   | **Per-intron** matrix (272191 features)             | —              |
| `sj_counts.1pz`       | Per-splice-junction counts                          | —              |
| `splice_psi.1pz`      | Per-event PSI (alt splicing)                        | derivable from sj_counts + splice_events.tsv |
| `gene_counts_em.1pz`  | EM-rescued multi-mapper counts                      | —              |
| `mt_heteroplasmy.1pz` | Mitochondrial allele matrix                         | —              |
| `vdj_gene_usage.1pz`  | V/D/J gene usage                                    | —              |

## Next steps

See the [interop vignette][] for end-to-end examples of standard
Bioconductor / Seurat workflows (`scran` normalization, `scater`
QC, Seurat UMAP + clustering, `scVelo`-compatible velocity).

[interop vignette]: interop.html
