---
title: "Getting Started with the Singlet Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the Singlet Atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The `singlet` package gives you programmatic access to the Singlet
single-cell genomics atlas. You work with two simple things: **GEO
accessions** (such as `"GSE138867"` or `"GSM4120733"`) and self-contained
**`.singlet` bundles**. The package handles downloading, caching, decoding,
and assembling the data into the single-cell objects you already use.

The chunks below need network access and download real studies, so they are
not run when the vignette is built. Every one of them works as written
against the public atlas.

## Install

```{r install}
install.packages(c("remotes", "BiocManager"))
BiocManager::install(c("SingleCellExperiment", "SummarizedExperiment", "S4Vectors"))
remotes::install_github("Singlet-Bio/singlet", subdir = "r")
```

```{r setup}
library(singlet)
```

`load()`, `find()`, `find_load()` and `download()` are also available as
`singlet_load()`, `singlet_find()`, `singlet_find_load()` and
`singlet_download()`. The short names mask `base::load()` and
`utils::find()` once the package is attached, so this vignette uses the
`singlet_` names.

## Load a study by accession

Pass a GEO Series accession to `singlet_load()`. The corresponding `.singlet`
bundle is downloaded from <https://data.singlet.bio>, cached locally, and
assembled into a `SingleCellExperiment`.

```{r load-one}
sce <- singlet_load("GSE138867")
sce
table(sce$gsm_id)
```

Subsequent calls reuse the cached bundle instead of re-downloading. The
cache lives in `tools::R_user_dir("singlet", "cache")` by default; override
it with the `cache_dir` argument or the `SINGLET_CACHE_DIR` environment
variable.

## Load a single sample

A GEO Sample accession works too. `singlet_load()` looks up the study it
belongs to, downloads that study's bundle, and keeps only the sample's cells.

```{r load-gsm}
sce <- singlet_load("GSM4120733")
unique(sce$gsm_id)
```

## Combine several studies

Give `singlet_load()` a vector of accessions (or local bundle paths, or a
mix) and it returns one combined object aligned on the shared gene axis.

```{r load-many}
sce <- singlet_load(c("GSE138867", "GSE146974"))
table(sce$gsm_id)
```

Samples with no usable cells (for example an empty count matrix left by a
failed pipeline run) are skipped with a warning instead of failing the load.
The metadata records which ones and why:

```{r skipped}
S4Vectors::metadata(sce)$skipped_samples
```

## Search with natural language

`singlet_find()` turns a free-text description into matching accessions.

```{r find}
hits <- singlet_find("T cells from pediatric AML")
hits

# One accession per sample instead of per study
singlet_find("human PBMC from smokers", level = "gsm", limit = 10)

# Search and load in a single step
sce <- singlet_find_load("microglia from Alzheimer's brain", limit = 5)
```

## Convert to Seurat

Ask for a `Seurat` object directly with `as = "seurat"` (this needs
`install.packages("Seurat")`).

```{r seurat}
obj <- singlet_load("GSE138867", as = "seurat")
obj

# Per-sample metadata is carried into the object's cell metadata
head(obj[[]])
```

## Other modalities

Bundles hold more than gene counts: splicing, mitochondrial heteroplasmy,
donor assignments, non-host reads and per-cell QC tables, depending on the
study and pipeline version. Download the bundle, check what a sample has, and
read what you need.

```{r modalities}
path <- singlet_download("GSE128639")
singlet_modalities(path)

gsm <- "GSM3681518"
if (singlet_has(path, "mt_heteroplasmy", gsm)) {
  het <- singlet_read(path, gsm, "mt_heteroplasmy")
  dim(het)
}
if (singlet_has(path, "cell_qc", gsm)) {
  head(singlet_read(path, gsm, "cell_qc"))
}

# One sample's raw counts, with spliced and unspliced assays
if (singlet_has(path, "exon_counts", gsm)) {
  raw <- singlet_raw_counts(path, gsm)
  SummarizedExperiment::assayNames(raw)
}
```

## Working with local bundles

If you already have a `.singlet` file on disk, read it directly with
`read_singlet()` (or just pass the path to `singlet_load()`).

```{r local}
path <- singlet_download("GSE138867")
sce <- read_singlet(path)
```

## Where the data comes from

Each cell carries its source sample in `colData(sce)$gsm_id`, alongside
tissue, cell type, disease, protocol, and other study metadata when
available. The parsed bundle manifest and study metadata are attached to
`metadata(sce)` for full provenance.

```{r provenance}
SummarizedExperiment::colData(sce)
names(S4Vectors::metadata(sce))
```
