---
title: "Getting Started with the Singlet Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the Singlet Atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The `singlet` package gives you programmatic access to the Singlet
single-cell genomics atlas. You work with two simple things: **GEO
accessions** (such as `"GSE149298"`) and self-contained **`.singlet`
bundles**. The package handles downloading, caching, decoding, and
assembling the data into the single-cell objects you already use.

```{r setup}
library(singlet)
```

## Load a study by accession

Pass a GEO Series accession to `load()`. The corresponding `.singlet`
bundle is downloaded from <https://singlet.bio>, cached locally, and
assembled into a `SingleCellExperiment`.

```{r load-one}
sce <- load("GSE149298")
sce
```

Subsequent calls reuse the cached bundle instead of re-downloading. The
cache lives in `tools::R_user_dir("singlet", "cache")` by default; override
it with the `cache_dir` argument or the `SINGLET_CACHE_DIR` environment
variable.

## Combine several studies

Give `load()` a vector of accessions (or local bundle paths, or a mix) and
it returns one combined object aligned on the shared gene axis.

```{r load-many}
sce <- load(c("GSE149298", "GSE116256"))
table(sce$gsm_id)
```

## Search with natural language

`find()` turns a free-text description into matching accessions.

```{r find}
hits <- find("T cells from pediatric AML", level = "gse")
hits

# Search and load in a single step
sce <- find_load("microglia from Alzheimer's brain", limit = 5)
```

## Convert to Seurat

Ask `load()` for a `Seurat` object directly with `as = "seurat"`.

```{r seurat}
obj <- load("GSE149298", as = "seurat")
obj

# Per-sample metadata is carried into the object's cell metadata
head(obj[[]])
```

## Working with local bundles

If you already have a `.singlet` file on disk, read it directly with
`read_singlet()` (or just pass the path to `load()`).

```{r local}
sce <- read_singlet("/data/GSE149298.singlet")
```

## Where the data comes from

Each cell carries its source sample in `colData(sce)$gsm_id`, alongside
tissue, cell type, disease, protocol, and other study metadata when
available. The parsed bundle manifest and study metadata are attached to
`metadata(sce)` for full provenance.

```{r provenance}
SummarizedExperiment::colData(sce)
names(SummarizedExperiment::metadata(sce))
```
