---
title: "Getting Started with singlepress"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with singlepress}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**singlepress** reads and writes `.1pz` sparse matrix files — a compressed format
achieving 13× compression with 4000+ MB/s decode speed. This vignette covers the
core R API: writing, reading, lazy access, indexing, subsetting, and summary statistics.

## Writing a .1pz file

Any `dgCMatrix` (sparse column-compressed matrix from the **Matrix** package) can be
written to `.1pz`:

```{r write-basic}
library(singlepress)
library(Matrix)

# Create a sparse matrix (30000 genes × 10000 cells)
set.seed(42)
mat <- rsparsematrix(30000, 10000, density = 0.05)
mat@x <- abs(mat@x)  # .1pz stores non-negative values
rownames(mat) <- paste0("gene_", seq_len(nrow(mat)))
colnames(mat) <- paste0("cell_", seq_len(ncol(mat)))

write_1pz(mat, "counts.1pz")
```

## Reading a .1pz file

```{r read}
mat <- read_1pz("counts.1pz", num_threads = 4)
dim(mat)            # [1] 30000 10000
rownames(mat)[1:3]  # "gene_1" "gene_2" "gene_3"
colnames(mat)[1:3]  # "cell_1" "cell_2" "cell_3"
```

Metadata attached as attributes:

```{r metadata}
attr(mat, "colsums")[1:5]        # pre-computed column sums
attr(mat, "obs")                 # cell metadata (data.frame)
attr(mat, "var")                 # gene metadata (data.frame)
attr(mat, "uns")                 # study metadata (named character)
```

## The OnePZ file handle

`OnePZ` is a lazy S4 class that reads only the file header — no matrix decompression
until you request data:
  
```{r onepz}
opz <- OnePZ("counts.1pz")
opz
# OnePZ: counts.1pz
#   Shape:   30,000 features x 10,000 cells
#   ...

dim(opz)     # [1] 30000 10000
nrow(opz)    # 30000
ncol(opz)    # 10000
n_obs(opz)   # 10000 (AnnData convention)
n_vars(opz)  # 30000 (AnnData convention)
```

### Indexing

```{r indexing}
# Integer slicing
sub <- opz[1:100, 1:500]

# By gene name
sub <- opz[c("gene_1", "gene_5", "gene_10"), ]

# First/last cells
head(opz, n = 10)
tail(opz, n = 10)

# Transpose
t_mat <- t(opz)  # cells × genes
```

### Type coercion

```{r coerce}
mat    <- as(opz, "dgCMatrix")     # sparse
dense  <- as(opz, "matrix")        # dense
df     <- as(opz, "data.frame")    # data.frame
```

## Summary Statistics

```{r stats}
# Non-zeros per column (cell)
npc <- nnz_per_col("counts.1pz")

# Non-zeros per row (feature)
npr <- nnz_per_row("counts.1pz")

# Row sums
rs <- rowsums_1pz("counts.1pz")

# Quick file description
desc <- describe_1pz("counts.1pz")
desc$shape     # c(30000, 10000)
desc$density   # 0.05
desc$file_size # bytes on disk
```

## Dataset Operations

### Concatenation

```{r concat}
# Column-bind (combine cells from multiple samples)
cbind_1pz("sample_A.1pz", "sample_B.1pz",
          output = "combined.1pz")

# Row-bind (combine features)
rbind_1pz("genes_A.1pz", "genes_B.1pz",
          output = "all_genes.1pz")
```

### Subsetting

```{r subset}
# By integer range
subset_1pz("counts.1pz", "sub.1pz",
           rows = 1:100, cols = 1:500)

# By gene name
subset_1pz("counts.1pz", "markers.1pz",
           rows = c("gene_1", "gene_5", "gene_10"))
```

### Random sampling

```{r sample}
# Sample 1000 cells
sample_1pz("counts.1pz", "sample.1pz", n = 1000, seed = 42)

# Sample 10%
sample_1pz("counts.1pz", "sample.1pz", fraction = 0.1)
```

## Log-Normalization

The `.1pz` format stores precomputed column sums, enabling on-the-fly
Seurat-style `LogNormalize` during reads — no separate normalization step:

```{r lognorm}
# Method 1: read_1pz with normalize = TRUE
mat_norm <- read_1pz("counts.1pz", normalize = TRUE, scale = 10000)

# Method 2: OnePZ handle in normalized mode
opz <- OnePZ("counts.1pz", normalize = TRUE)
opz[1:100, 1:500]   # returns log-normalized submatrix

# Method 3: create a normalized view from existing handle
opz <- OnePZ("counts.1pz")
npz <- normalized(opz)
as(npz, "dgCMatrix")  # full log-normalized matrix
```

The formula is `log1p(x / colsum * scale)`, where `scale` defaults to 10000.

For manual normalization of already-loaded data:

```{r lognorm_manual}
mat <- read_1pz("counts.1pz")
cs <- attr(mat, "colsums")
norm <- lognorm(mat, cs, scale = 10000)
```

## File Inspection

```{r inspect}
info <- info_1pz("counts.1pz")
info$m            # 30000 rows (genes)
info$n            # 10000 cols (cells)
info$nnz          # number of non-zeros
info$has_obs_var  # TRUE if obs/var DataFrames stored

# CRC32 integrity validation
result <- validate_1pz("counts.1pz")
result$valid  # TRUE
```
