---
title: "SingleCellExperiment Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SingleCellExperiment Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette shows how singlepress integrates with the Bioconductor
**SingleCellExperiment** (SCE) ecosystem. Read `.1pz` files into SCE objects
with colData, rowData, and metadata preserved.

## Reading .1pz → SingleCellExperiment

```{r read-sce}
library(singlepress)
library(SingleCellExperiment)

sce <- read_1pz_sce("counts.1pz", num_threads = 4)
sce
# class: SingleCellExperiment
# dim: 30000 10000
# assays(1): counts
# rowData names(2): gene_name highly_variable
# colData names(4): cell_type tissue donor total_counts
```

### What gets imported

| .1pz section | SCE location |
|---|---|
| Count matrix | `assay(sce, "counts")` / `counts(sce)` |
| Gene names (rownames) | `rownames(sce)` |
| Cell barcodes (colnames) | `colnames(sce)` |
| Column sums | `colData(sce)$total_counts` |
| obs DataFrame | `colData(sce)` |
| var DataFrame | `rowData(sce)` |
| uns key-value pairs | `metadata(sce)` |

```{r inspect-sce}
# Count matrix
counts(sce)[1:5, 1:5]

# Cell metadata
head(colData(sce))

# Gene metadata
head(rowData(sce))

# Study-level metadata
metadata(sce)
```

## Using OnePZ Coercion

If SingleCellExperiment is installed, use `as()` on an OnePZ handle:

```{r coerce-sce}
opz <- OnePZ("counts.1pz")
sce <- as(opz, "SingleCellExperiment")
```

## Bioconductor Workflow

The imported SCE works with all Bioconductor tools:

```{r bioc-workflow}
library(scater)
library(scran)

# Quality control
sce <- addPerCellQCMetrics(sce)
sce <- addPerFeatureQCMetrics(sce)

# Filtering
keep_cells <- sce$sum > 500 & sce$detected > 200
keep_genes <- rowData(sce)$detected > 10
sce <- sce[keep_genes, keep_cells]

# Normalization
clusters <- quickCluster(sce)
sce <- computeSumFactors(sce, clusters = clusters)
sce <- logNormCounts(sce)

# Feature selection
dec <- modelGeneVar(sce)
top_hvgs <- getTopHVGs(dec, n = 2000)

# Dimensionality reduction
sce <- runPCA(sce, subset_row = top_hvgs)
sce <- runUMAP(sce, dimred = "PCA")

plotUMAP(sce, colour_by = "cell_type")
```

## Using Stored Column Sums

The `.1pz` format stores pre-computed column sums. These are imported into
`colData(sce)$total_counts`:

```{r colsums-sce}
# Already available after read_1pz_sce
head(colData(sce)$total_counts)

# Use for quick size factor computation
sizeFactors(sce) <- colData(sce)$total_counts / mean(colData(sce)$total_counts)
```

## Writing SCE → .1pz

```{r write-sce}
write_1pz_sce(sce, "output.1pz", assay_name = "counts")
```

## Subsetting Before Import

For large datasets, filter the `.1pz` file before creating the SCE:

```{r subset-sce}
# Quick exploration: sample 2000 cells
sample_1pz("atlas.1pz", "sample.1pz", n = 2000, seed = 42)
sce_sample <- read_1pz_sce("sample.1pz")

# Select marker genes only
subset_1pz("atlas.1pz", "markers.1pz",
           rows = c("CD3D", "CD3E", "CD4", "CD8A",
                    "CD19", "MS4A1", "CD14", "FCGR3A"))
sce_markers <- read_1pz_sce("markers.1pz")
```

## Multi-Sample Integration

```{r multi-sample-sce}
# Combine at the .1pz level
cbind_1pz("donor_1.1pz", "donor_2.1pz", "donor_3.1pz",
          output = "merged.1pz")
sce <- read_1pz_sce("merged.1pz")

# Or combine SCE objects directly
sce1 <- read_1pz_sce("donor_1.1pz")
sce2 <- read_1pz_sce("donor_2.1pz")
sce_merged <- cbind(sce1, sce2)
```

## Comparison with Other Formats

| Format | Read time (10K cells) | File size | Random access |
|---|---|---|---|
| `.1pz` | ~0.1s | 45 MB | Yes (column/row ranges) |
| `.h5ad` | ~0.5s | 120 MB | Limited |
| `.rds` (dgCMatrix) | ~2s | 200 MB | No |
| `.mtx.gz` | ~3s | 180 MB | No |

The `.1pz` format is particularly advantageous for:
- Fast partial reads (only decompress needed columns)
- Pre-computed column sums (skip library size calculation)
- Cross-language compatibility (same file works in Python and R)
