---
title: "Seurat Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seurat Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates the Seurat integration in singlepress. You can read `.1pz`
files directly into Seurat objects (with metadata preserved), run standard Seurat
workflows, and write results back to `.1pz`.

## Reading .1pz → Seurat

```{r read-seurat}
library(singlepress)
library(Seurat)

obj <- read_1pz_seurat("counts.1pz", assay_name = "RNA")
obj
# An object of class Seurat
# 30000 features across 10000 samples within 1 assay
# Active assay: RNA (30000 features, 0 variable features)
```

### What gets imported

| .1pz section | Seurat location |
|---|---|
| Count matrix | `GetAssayData(obj, layer = "counts")` |
| Gene names (rownames) | `rownames(obj)` |
| Cell barcodes (colnames) | `colnames(obj)` |
| Column sums | `obj$total_counts` |
| obs DataFrame | Merged into `obj@meta.data` |
| uns key-value pairs | `Seurat::Misc(obj)` |

```{r inspect-seurat}
# Cell metadata from .1pz obs
head(obj@meta.data)

# Pre-computed library size
head(obj$total_counts)

# Study metadata
Seurat::Misc(obj)
```

## Using OnePZ Coercion

If Seurat is installed, you can use `as()` directly on an OnePZ handle:

```{r coerce-seurat}
opz <- OnePZ("counts.1pz")
obj <- as(opz, "Seurat")
```

## Standard Seurat Workflow

Once imported, the object works exactly like any Seurat object:

```{r workflow}
# Normalization (or use singlepress lognorm for pre-computed colsums)
obj <- NormalizeData(obj)

# Variable features
obj <- FindVariableFeatures(obj, nfeatures = 2000)

# Scale and PCA
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 30)

# Clustering
obj <- FindNeighbors(obj, dims = 1:20)
obj <- FindClusters(obj, resolution = 0.5)

# UMAP
obj <- RunUMAP(obj, dims = 1:20)
DimPlot(obj, group.by = "cell_type")
```

## Using Stored Column Sums

The `.1pz` format pre-computes column sums during encoding. Use these for
normalization instead of recalculating:

```{r stored-colsums}
mat <- read_1pz("counts.1pz")
cs <- attr(mat, "colsums")

# Fast log-normalization
norm <- lognorm(mat, cs, scale = 10000)

# Or use total_counts in the Seurat object
obj$nCount_RNA <- obj$total_counts
```

## Writing Seurat → .1pz

Save a Seurat object back to `.1pz`:

```{r write-seurat}
write_1pz_seurat(obj, "output.1pz", assay_name = "RNA")
```

This extracts the count matrix and writes it with rownames and colnames preserved.

## Subsetting Before Import

For large datasets, subset the `.1pz` file before importing to Seurat:

```{r subset-then-import}
# Sample 5000 cells first
sample_1pz("large_dataset.1pz", "sampled.1pz", n = 5000, seed = 42)

# Import only marker genes
subset_1pz("large_dataset.1pz", "markers.1pz",
           rows = c("CD3D", "CD3E", "CD4", "CD8A", "CD19", "MS4A1"))

# Then read into Seurat
obj <- read_1pz_seurat("sampled.1pz")
```

## Multiple Samples

Combine multiple `.1pz` files into one Seurat object:

```{r multi-sample}
# Method 1: cbind → single read
cbind_1pz("sample_1.1pz", "sample_2.1pz", "sample_3.1pz",
          output = "merged.1pz")
obj <- read_1pz_seurat("merged.1pz")

# Method 2: Read separately, merge in Seurat
obj1 <- read_1pz_seurat("sample_1.1pz")
obj2 <- read_1pz_seurat("sample_2.1pz")
obj <- merge(obj1, obj2)
```
