---
title: "SinglePress (.1pz) — R Companion"
author: "Zachary DeBruine"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Installation

Install the `singlepress` R package from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("Singlet-Bio/singlepress", subdir = "R")
```

```{r load}
library(singlepress)
```

## Read a 1pz File

This assumes `pbmc3k.1pz` was created by the Python companion notebook.
Adjust the path to match your local setup.

```{r read}
pz_path <- "pbmc3k.1pz"  # adjust path as needed
opz <- OnePZ(pz_path)
opz
```

## Inspect Dimensions

```{r dims}
cat(sprintf("Cells:    %d\n", ncol(opz)))
cat(sprintf("Genes:    %d\n", nrow(opz)))
cat(sprintf("Nonzeros: %s\n", format(describe_1pz(pz_path)$nnz, big.mark = ",")))
```

## Convert to dgCMatrix

The standard sparse matrix class used by Seurat, scran, and other
Bioconductor tools:

```{r convert}
mat <- as(opz, "dgCMatrix")
cat(sprintf("Class: %s\n", class(mat)))
cat(sprintf("Dimensions: %d × %d\n", nrow(mat), ncol(mat)))
cat(sprintf("Density: %.4f%%\n", 100 * length(mat@x) / prod(dim(mat))))
```

## Column Subsetting (Lazy)

SinglePress supports lazy column slicing — only the requested cells are
decompressed:

```{r subset}
first100 <- opz[, 1:100]
cat(sprintf("Subset: %d genes × %d cells\n", nrow(first100), ncol(first100)))
```

## Benchmark: 1pz vs MatrixMarket

Compare read time for the same dataset stored as Market Matrix (.mtx)
vs SinglePress (.1pz):

```{r benchmark, eval=FALSE}
# Requires Matrix package and the original MTX directory
library(Matrix)

mtx_path <- "filtered_gene_bc_matrices/hg19/matrix.mtx"

t_mtx <- system.time(readMM(mtx_path))["elapsed"]
t_1pz <- system.time(as(OnePZ(pz_path), "dgCMatrix"))["elapsed"]

cat(sprintf("MatrixMarket read: %.3f s\n", t_mtx))
cat(sprintf("SinglePress read:  %.3f s\n", t_1pz))
cat(sprintf("Speedup: %.1f×\n", t_mtx / t_1pz))
```

## Session Info

```{r session}
sessionInfo()
```
