---
title: "GPU Analysis with singlet"
author: "Zachary DeBruine"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{GPU Analysis with singlet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # GPU required; set eval=TRUE in an interactive session
)
```

# Overview

The **singlet** R package includes GPU-accelerated analysis when CUDA 12+ is
available. It reads `.1pz` count matrices and runs GPU-native:

- Log-normalization (`lognorm`)
- Highly-variable gene selection (`hvg`)
- PCA via SVD backends (`run_pca`)
- NMF via MU/CD/Cholesky (`run_nmf`)

An optional reticulate bridge provides Harmony batch correction and UMAP via
`rapids-singlecell`.

# Requirements

- R >= 4.2
- CUDA Toolkit >= 12.0
- GPU with compute capability >= 7.0 (Volta, Turing, Ampere, or Hopper)

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("SingleCellExperiment", "SummarizedExperiment"))

remotes::install_github("Singlet-Bio/singlet", subdir = "r")
```

Before installing with GPU support, ensure `nvcc` is on your PATH:

```bash
export PATH=/usr/local/cuda/bin:$PATH
```

# Quick start

```{r quickstart}
library(singlet)

# Load a .1pz file
sce <- read_1pz("gene_counts.1pz", as = "sce")
dim(sce)

# Check GPU availability
singlet::has_gpu()

# Log-normalization (result in assays(sce)$logcounts)
sce <- lognorm(sce, target_sum = 1e4)

# Select 2000 highly variable genes
sce <- hvg(sce, n_top = 2000, flavor = "seurat_v3")

# PCA (50 components)
sce <- run_pca(sce, n_pcs = 50)

# NMF (20 factors, MSE loss)
sce <- run_nmf(sce, n_factors = 20, loss = "MSE")

# Access embeddings
head(reducedDim(sce, "PCA"))
```

# Reticulate bridge (optional)

If `reticulate` and the `singlet` Python package are installed:

```{r harmony}
# Add a batch covariate
colData(sce)$sample_id <- "sample_1"

# Harmony batch correction
sce <- rapids_harmony(sce, batch_key = "sample_id")

# UMAP
sce <- rapids_umap(sce, basis = "X_pca_harmony")
plot(reducedDim(sce, "UMAP"), pch = ".", main = "UMAP")
```

# Performance notes

- R does not have a native GPU sparse type. Every kernel call incurs an O(nnz)
  host copy on input (dgCMatrix → pinned host CSC) and output.
- For a typical 10k-cell sample this is approximately 100–200 ms per call.
- Users who require full GPU residency between operations should use the Python
  wrapper (`singlet[gpu]`, cupy zero-copy path).

# Session info

```{r sessioninfo}
sessionInfo()
```
